# Methods

## VSEM model
 * very simple toy model. not realistic but simplified form with and similar structure to many ecosystem models. 

Photosynthesis equation

\begin{align}
NPP &= PAR \times LUE \times (1 - \exp^{(-KEXT \times C_v)})\\
\end{align}

 * PAR Photosynthetically active radiation
 * LUE Light use efficiency of NPP (Ra implicit)
 * KEXT Beer's law light extinction coeff
 * $C_v$ Vegetation carbon

Carbon pool state equations

\begin{align}
\frac{dC_v}{dt}  &= A_v \times NPP &- \frac{C_v}{\tau_v} \\
\frac{dC_r}{dt}  &= (1.0-A_v) \times NPP &- \frac{C_r}{\tau_r}\\
\frac{dC_s}{dt}  &= \frac{C_r}{\tau_r} + \frac{C_v}{\tau_v} &- \frac{C_s}{\tau_s}
\end{align}
 - $C_v$, $C_r$ and $C_s$ : Carbon in vegetation, root and soil pools

## Bayesian Calibration
 * refer to forthcoming TG13 paper
 * R package used BayesianTools
 * DREAMzs algorithm

## Idealised experiments with virtual data from VSEM
```{r child = 'TG15-virtualData.Rmd'}
```
 * advantage we know what the truth is.
 * balanced versus unbalanced 
 * perfect model and model with known error
 * data with and without known bias
 * include additive and multiplicative parameters to represent structural errors in the model and systematic biases in the data. 

```{r include=FALSE}
## Load older versions of BayesianTools R routines needed in TG15
source("TG15-BayesianToolsOld.R")
source("helperFunctions.R")
```


