# Methods

## VSEM model

Here we present the Very Simple Ecosystem Model (VSEM). The model was created to help illustrate the
main ideas that we present here. The model was designed to be very simple rather then realistic, but
yet resemble many typical, but more complicated, process-based ecosystem models (PBMs) that are
commonly used in carbon growth type ecosystem modelling.

In essence, the model determines the accumulation of carbon in the plant and soil from the growth of
the plant via photosynthesis and senescence to the soil which respires carbon back to the
atmosphere. The timestep of the VSEM is daily. 

### VSEM input data: Photosythetically active radiation (PAR)

The VSEM requires only one input dataset to drive the model namely daily PAR. 

Since we are interested in virtual experiments here we generate the PAR input data using an
sinusoidal function.

\begin{align}
PAR &= (\lvert \sin ( Days /365 \times \pi)  + \epsilon \rvert) \times 10 \\
\end{align}

 * PAR Photosynthetically active radiation
 * $\epsilon$ Gaussian noise added
 * Days number of days

### Photosynthesis equation

The model calculates Gross Primary Productivity (GPP) using a very simple light-use efficiency (LUE)
formulation multiplied by light interception. Light interception is calculated via Beer's law with a
constant light extinction coefficient operating on Leaf Area Index (LAI).  A parameter (GAMMA)
determines the fraction of GPP that is autotrophic respiration, giving the Net Primary Productivity (NPP).

\begin{align}
NPP &= PAR \times LUE \times (1 - \exp^{(-KEXT \times C_v)})\\
\end{align}

 * PAR Photosynthetically active radiation (MJ $m^{-2}$ $day^{-1}$)
 * LUE Light use efficiency of NPP (Ra implicit)
 * KEXT Beer's law light extinction coeff
 * $C_v$ Vegetation carbon

### Carbon pool state equations

There are three state equations [@Gill] representing the change in time of vegetation ($C_v$), root ($C_r$)
and soil ($C_s$) carbon pools. The Net Primary Productivity (NPP) is allocated to above (vegetation)
and below(root) ground carbon pools via a fixed allocation fraction. Carbon is lost from the plant
pools to a single soil pool via fixed vegetation and root turnover rates. Heterotropic respiration
in the soil is determined via a soil turnover rate.

\begin{align}
\frac{dC_v}{dt}  &= A_v \times NPP &- \frac{C_v}{\tau_v} \\
\frac{dC_r}{dt}  &= (1.0-A_v) \times NPP &- \frac{C_r}{\tau_r}\\
\frac{dC_s}{dt}  &= \frac{C_r}{\tau_r} + \frac{C_v}{\tau_v} &- \frac{C_s}{\tau_s}
\end{align}

## Bayesian Calibration
 * refer to forthcoming TG13 paper
 * R package used BayesianTools
 * DREAMzs algorithm

## Idealised experiments with virtual data from VSEM
```{r child = 'TG15-virtualData.Rmd'}
```

### Perfect model

A central theme that we consider here is the significance of a perfect model structure with all the
processes modelled perfectly. The only way to ensure a perfect model is to take the output from the
VSEM and consider this as virtual data in the BC. Gaussian noise is added to the model output to
represent system varability that is not captured by the model (as is away the case) but crucially
can be represented perfectly by the likelihood function that we use in the BC. The observations are
for the full 2048 day length of the VSEM for NEE, vegetative carbon and soil carbon.

For the vegetaive carbon we create a sparse dataset to simulate having an imbalance between
observations available for vegetative carbon, soil carbon and NEE. The sparse dataset has six
observations for days 2, 404, 780, 1100, 1500 and 1840.

### Model with known structural error

To simulate a model with a known structural error we consider a situation where a major model
process/structure is unknown and therefore missing in the model. Here we remove the root pool
completely from the VSEM to simulate a major structural error. This is done by initialising the root
pool to zero and setting the root allocation fraction to zero so that all the NPP is now allocated
to the vegetation pool. This also of course shuts off any senescence from the root pool to the
soil. This gives the model a major structural error as we might have in a real situation whilst
being sufficiently simple that we can still interpret the influence of the error.

### Observational data with known bias

In addition to considering model structural error, we also wish to investigate the influence of
observations with biases since all observational data will to a greater or lesser extent contain
biases. Here we simulate data biases by multiplying the soil data by two to represent a considerable
multiplicative bias in the observations of soil carbon. 

## Modified Likelihood to represent structural errors in the model and systematic biases in the data. 

Present the inclusion of additive and multiplicative parameter bias parameters in the likelihood
here or leave until results section?


```{r include=FALSE}
## Load older versions of BayesianTools R routines needed in TG15
source("TG15-BayesianToolsOld.R")
source("helperFunctions.R")
```


